### 2013-07-22 docs/Jesse/Dissertation/DPGtheory.tex
updating diss files

\subsection{Discontinuous Petrov-Galerkin methods with optimal test functions}
\seclab{optimalTest} Petrov-Galerkin methods, in which the test space
differs from the trial space, have been explored for over 30 years,
beginning with the approximate symmetrization method of Barrett and
Morton~\cite{BARRETT01101981}. The idea was continued with the SUPG
method of Hughes, and the characteristic Petrov-Galerkin approach of
Demkowicz and Oden~\cite{Demkowicz1986188}, which introduced the
idea of tailoring the test space to change the norm in which a finite
element method would converge.

@@

Throughout this dissertation, we assume that the trial space $U$ and test space $V$ are real Hilbert spaces, and denote $U'$ and $V'$ as the respective topological dual spaces. Let $U_h \subset U$ and $V_h\subset V$ be finite dimensional subsets. We are interested in the following problem  
\begin{equation}
\eqnlab{variationEq}
\left\{
  \begin{array}{l}
    \text{Given } l \in V', \text{ find } u_h \in U_h  \text{ such that} \\ 
    b(u_h,v_h) = l(v_h), \quad \forall v_h\in V_h,
  \end{array}
  \right.
\end{equation}
where $b\LRp{\cdot,\cdot}: U \times V \to \mbb{R}$ is a continuous
bilinear form.  $U_h$ is chosen to be some trial space of approximating functions, but $V_h$ is as of yet unspecified.

Throughout the paper, we suppose the variational
problem \eqnref{variationEq} to be well-posed. In that case, we can
identify a unique operator $B:U\rightarrow V'$ such that
\[
\langle Bu,v\rangle_V \coloneqq b(u,v), \quad u\in U, v\in V
\]
with $\LRa{\cdot, \cdot}_V$ denoting the duality pairing between $V'$ and $V$, to obtain the operator form of the continuous variational problem
\begin{equation}
\eqnlab{dualEq}
Bu = l \quad \text{in } V'.
\end{equation}
In other words, we can represent the continuous form of our variational equation
\eqnref{variationEq} equivalently as the operator equation \eqnref{dualEq} with values in the
dual space $V'$.  This motivates us to consider the conditions under which the solution to \eqnref{variationEq} is the solution to the minimum residual problem in $V'$ 
\[
u_h = \underset{u_h\in U_h}{\arg\min}\, J(u_h),
\]
where $J(w)$ is defined for $w\in U$ as 
\[
J(w) = \frac{1}{2}\|Bw-l\|_{V'}^2 \coloneqq\frac{1}{2} \sup_{v\in V\setminus\{0\}} \frac{| b(w,v)-l(v)|^2}{\nor{v}_V^2}.
\]
For convenience in writing, we will abuse the notation $\sup_{v \in V}$ to denote $\sup_{v\in V\setminus\{0\}}$ for the remainder of the paper.

@@

We note that solving \eqnref{optvVar} on each element exactly is infeasible since it amounts to inverting the Riesz map $R_V$ exactly. Instead, optimal test functions are approximated using the standard Bubnov-Galerkin method on an ``enriched" subspace $\tilde{V} \subset V$ such that $\dim(\tilde{V}) > \dim(U_h)$ elementwise \cite{DPG1, DPG2}. In this paper, we assume the error in approximating the optimal test functions is negligible, and refer to the work in \cite{practicalDPG} for estimating the effects of approximation error on the performance of DPG.

@@

\subsection{Duality between trial and test norms (energy norm pairings)}
%\subsection{Abstract energy norm pairings}
\seclab{energyPair}
A clear property of the energy norm defined by \eqnref{energyNorm} is that the trial norm $\nor{\cdot}_{U,E}$ is induced by a given test norm. However, the reverse relationship holds as well; for any trial norm, the test norm that induces such a norm is recoverable through duality. We have a result, Lemma 2.5 in \cite{Bui-ThanhDemkowiczGhattas11a}: assuming, for simplicity, that the bilinear form $b(u,v)$ is definite\footnote{By definite, we mean that \begin{align*}b(u,v) &= 0, \forall v\in V \Rightarrow u = 0\\ b(u,v)&=0, \forall u\in U \Rightarrow v = 0,\end{align*}  which imply injectivity of the bilinear operator $B$ and its transpose $B'$, defined such that $\LRa{Bu,v} = \LRa{u,B'v}$.  These conditions imply solvability of the variational problem.}, given any norm $\nor{\cdot}_{U}$ on the trial space $U$, for $\varphi \in U$, we can represent $\nor{\varphi}_{U}$ via
\[
\nor{\varphi}_{U} = \sup_{v \in V}\frac{b\LRp{w,v}}{\nor{v}_{V,U}}.
\]
where $\nor{v}_{V,U}$ is defined through
\[
\nor{v}_{V,U} = \sup_{w \in U}\frac{b\LRp{w,v}}{\nor{w}_{U}}.
\]
%In other words, the norm $\nor{\cdot}_V$ in $V$ can be recovered using the energy norm $\nor{\cdot}_{U,E}$ in $U$, and vice versa.

@@

%From now on, unless otherwise stated, we will refer to  $\nor{\cdot}_{V,U,1}$, $\nor{\cdot}_{V,U,2}$, and $\nor{\cdot}_{V}$ as the test norms which induce $\nor{\cdot}_{U,1}$, $\nor{\cdot}_{U,2}$, and $\nor{\cdot}_{U,E}$ and vice versa, in the sense discussed above. 
From now on, unless otherwise stated, we will refer to $\nor{\cdot}_{V,U}$ as the test norm that induces a given norm $\nor{\cdot}_U$. Likewise, we will refer $\nor{\cdot}_{U,V}$ as the trial norm induced by a given test norm $\nor{\cdot}_V$. In this paper, for simplicity of exposition, we shall call a pair of norms in $U$ and $V$ that induce each other as an {\em energy norm pairing}.

\subsection{Discontinuous Petrov-Galerkin methods with the ultra-weak formulation}
\seclab{abstractUweak}

The naming of the discontinuous Petrov-Galerkin method refers to the fact that the method is a Petrov-Galerkin method, and that the test functions are specified to be discontinuous across element boundaries. There is no specification of the regularity of the trial space, and we stress that the idea of DPG is not inherently tied to a single variational formulation \cite{Bui-ThanhDemkowiczGhattas11a}. Additionally, Cohen, Dahmen and Welper simultaneously extend the minimum residual concept behind DPG to general variational settings and avoid the use of discontinuous test functions by formulating the minimum residual method as a saddle-point problem \cite{DahmenVariationalStabilization}.

In most of the DPG literature, however, the discontinuous Petrov-Galerkin method refers to the combination of the concept of locally computable optimal test functions in Section \secref{optimalTest} with the so-called ``ultra-weak formulation" \cite{DPG1,DPG2,DPG3,DPG4,DPGElas,DBLP:journals/procedia/NiemiCC11}. Unlike the previous two sections in which we studied the general equation \eqnref{variationEq} given by abstract bilinear and linear forms, we now consider a concrete instance of \eqnref{variationEq} resulting from an ultra-weak formulation for an abstract first-order system of PDEs $Au = f$. Additionally, from this section onwards, we will refer to DPG as the pairing of the ultra-weak variational formulation with the concept of locally computable optimal test functions.

We begin by partitioning the domain of interest $\Omega$ into $\Nel$ non-overlapping elements $K_j, j = 1,\hdots,\Nel$ such that $\Oh = \cup_{j=1}^\Nel K_j$ and $\overline{\Omega} = \overline{\Omega}_h$. Here, $h$ is defined as $h= \max_{j\in \LRc{1,\hdots,\Nel}}\text{diam}\LRp{K_j}$.  We denote the mesh ``skeleton" by $\Gh = \cup_{j=1}^\Nel \partial K_j$; the set of all faces/edges $e$, each of which comes with a normal vector ${n}_e$. The internal skeleton is then defined as $\Gamma^0_h = \Gh \setminus \partial \Omega$. If a face/edge $e \in \Gh$ is the intersection of $\partial K_i$ and $\partial K_j$, $i \ne j$, we define the following jumps:
\[
\jump{v} = \text{sgn} \LRp{{n}^-}v^- + \text{sgn} \LRp{{n}^+}v^+, \quad
\jump{\tau \cdot n} = {n}^-\cdot \tau^- + {n}^+\cdot\tau^+,
\]
where
\[
\text{sgn}\LRp{{n}^{\pm}} =
\left\{
\begin{array}{ll}
1 & \text{if } {n}^{\pm} = {n}_e \\
-1 & \text{if } {n}^\pm = -{n}_e
\end{array}
\right..
\]
For $e$ belonging to the domain boundary $\partial \Omega$, we define
\[
\jump{v} = v, \quad
\jump{\tau \cdot n} = {n}_e\cdot \tau.
\]
Note that we allow arbitrariness in assigning ``-'' and ``+'' quantities to the adjacent elements $K_i$ and $K_j$.
%For the rest of the paper, we will use the same notation for both a function and its trace (if it is well-defined) when there is no ambiguity.

The ultra-weak formulation for $Au = f$ on $\Oh$, ignoring boundary
conditions for now, reads
\begin{equation}
\eqnlab{uweak}
b\left(\left(u, \widehat{u}\right),v\right) \coloneqq \langle \widehat{u}, \jump{v}
\rangle_{\Gh} - (u,A_h^*v)_{\Oh}= \LRp{f,v}_{\Oh},
\end{equation}
where we have denoted $\LRa{\cdot,\cdot}_{\Gh}$ as the duality
pairing on $\Gh$, $\LRp{\cdot,\cdot}_{\Oh}$ the $L^2$-inner
product over $\Oh$, and $A_h^*$ the formal adjoint resulting from
element-wise integration by parts.  Occasionally, for simplicity in
writing, we will ignore the subscripts in the duality pairing and
$L^2$-inner product if they are $\Gh$ and $\Oh$. Both the
inner product and formal adjoint are understood to be taken
element-wise. Using the ultra-weak formulation, the regularity
requirement on solution variable $u$ is relaxed, that is, $u$ is now
square integrable for the ultra-weak formulation \eqnref{uweak} to be
meaningful, instead of being (weakly) differentiable.  The trade-off
is that $u$ does not admit a trace on $\Gh$ even though it did
originally. Consequently, we need to introduce an additional new
``trace'' variable $\widehat{u}$ in \eqnref{uweak}, that is defined only on
$\Gh$.

@@

%\subsection{Optimal and quasi-optimal test norms}
\subsection{A canonical energy norm pairing for ultra-weak formulation}

@@

Using variants of the quasi-optimal test norm, numerical results show that the DPG method appears to provide a ``pollution-free" method without phase error for the Helmholtz equation \cite{DPG4}, and analysis of the pollution-free nature of DPG is currently under investigation. Similar results have also been obtained in the context of elasticity \cite{DPGElas} and the linear Stokes equations \cite{Camellia}. On the theoretical side, the quasi-optimal test norm has been shown to yield a well-posed DPG methodology for the Poisson and convection-diffusion equations \cite{analysisDPG}. More recently, this theory has been generalized to show the well-posedness of DPG for the large class of PDEs of Friedrichs' type \cite{Bui-ThanhDemkowiczGhattas11b}.

>>>>

\section{Discontinuous Petrov-Galerkin methods with optimal test functions}
\seclab{optimalTest} Petrov-Galerkin methods, in which the test space
differs from the trial space, have been explored for over 30 years,
beginning with the approximate symmetrization method of Barrett and
Morton~\cite{BARRETT01101981}. The idea was continued with the SUPG
method of Hughes, and the characteristic Petrov-Galerkin approach of
Demkowicz and Oden~\cite{Demkowicz1986188}, which introduced the
idea of tailoring the test space to change the norm in which a finite
element method would converge.

@@

Throughout this dissertation, we assume that the trial space $U$ and test space $V$ are real Hilbert spaces, and denote $U'$ and $V'$ as the respective topological dual spaces. Let $U_h \subset U$ and $V_h\subset V$ be finite dimensional subspaces. We are interested in the following problem  
\begin{equation}
\eqnlab{variationEq}
\left\{
  \begin{array}{l}
    \text{Given } l \in V', \text{ find } u_h \in U_h  \text{ such that} \\ 
    b(u_h,v_h) = l(v_h), \quad \forall v_h\in V_h,
  \end{array}
  \right.
\end{equation}
where $b\LRp{\cdot,\cdot}: U \times V \to \mbb{R}$ is a continuous
bilinear form.  $U_h$ is chosen to be some trial space of approximating functions, but $V_h$ is as of yet unspecified.

Throughout the dissertation, we suppose the variational
problem \eqnref{variationEq} to be well-posed. In that case, we can
identify a unique operator $B:U\rightarrow V'$ such that
\[
\langle Bu,v\rangle_V \coloneqq b(u,v), \quad u\in U, v\in V
\]
with $\LRa{\cdot, \cdot}_V$ denoting the duality pairing between $V'$ and $V$, to obtain the operator form of the continuous variational problem
\begin{equation}
\eqnlab{dualEq}
Bu = l \quad \text{in } V'.
\end{equation}
In other words, we can represent the continuous form of our variational equation
\eqnref{variationEq} equivalently as the operator equation \eqnref{dualEq} with values in the
dual space $V'$.  This motivates us to consider the conditions under which the solution to \eqnref{variationEq} is the solution to the minimum residual problem in $V'$ 
\[
u_h = \underset{u_h\in U_h}{\arg\min}\, J(u_h),
\]
where $J(w)$ is defined for $w\in U$ as 
\[
J(w) = \frac{1}{2}\|Bw-l\|_{V'}^2 \coloneqq\frac{1}{2} \sup_{v\in V\setminus\{0\}} \frac{| b(w,v)-l(v)|^2}{\nor{v}_V^2}.
\]
For convenience in writing, we will abuse the notation $\sup_{v \in V}$ to denote $\sup_{v\in V\setminus\{0\}}$ for the remainder of the dissertation.

@@

We note that solving \eqnref{optvVar} on each element exactly is still infeasible since it amounts to inverting the Riesz map $R_V$ exactly. Instead, optimal test functions are approximated using the standard Bubnov-Galerkin method on an ``enriched" subspace $\tilde{V} \subset V$ such that $\dim(\tilde{V}) > \dim(U_h)$ elementwise \cite{DPG1, DPG2}. In this dissertation, we assume the error in approximating the optimal test functions is negligible, and refer to the work in \cite{practicalDPG} for estimating the effects of approximation error on the performance of DPG.

@@

\section{Duality between trial and test norms (energy norm pairings)}

\seclab{energyPair}
A clear property of the energy norm defined by \eqnref{energyNorm} is that the trial norm $\nor{\cdot}_{U,E}$ is induced by a given test norm. However, the reverse relationship holds as well; for any trial norm, the test norm that induces such a norm is recoverable through duality. We have a result, Lemma 2.5 in \cite{Bui-ThanhDemkowiczGhattas11a}: assuming, for simplicity, that the bilinear form $b(u,v)$ is definite\footnote{By definite, we mean that \begin{align*}b(u,v) &= 0, \forall v\in V \Rightarrow u = 0\\ b(u,v)&=0, \forall u\in U \Rightarrow v = 0,\end{align*}  which imply injectivity of the bilinear operator $B$ and its transpose $B'$, defined such that $\LRa{Bu,v} = \LRa{u,B'v}$.  These conditions imply solvability of the variational problem.}, given any norm $\nor{\cdot}_{U}$ on the trial space $U$, for $\varphi \in U$, we can represent $\nor{\varphi}_{U}$ via
\[
\nor{\varphi}_{U} = \sup_{v \in V}\frac{b\LRp{\varphi,v}}{\nor{v}_{V,U}},
\]
where $\nor{v}_{V,U}$ is defined through
\[
\nor{v}_{V,U} = \sup_{w \in U}\frac{b\LRp{w,v}}{\nor{w}_{U}}.
\]
%In other words, the norm $\nor{\cdot}_V$ in $V$ can be recovered using the energy norm $\nor{\cdot}_{U,E}$ in $U$, and vice versa.

@@

%From now on, unless otherwise stated, we will refer to  $\nor{\cdot}_{V,U,1}$, $\nor{\cdot}_{V,U,2}$, and $\nor{\cdot}_{V}$ as the test norms which induce $\nor{\cdot}_{U,1}$, $\nor{\cdot}_{U,2}$, and $\nor{\cdot}_{U,E}$ and vice versa, in the sense discussed above. 
From now on, unless otherwise stated, we will refer to $\nor{\cdot}_{V,U}$ as the test norm that induces a given norm $\nor{\cdot}_U$. Likewise, we will refer $\nor{\cdot}_{U,V}$ as the trial norm induced by a given test norm $\nor{\cdot}_V$. In this dissertation, for simplicity of exposition, we shall call a pair of norms in $U$ and $V$ that induce each other as an {\em energy norm pairing}.

\section{Discontinuous Petrov-Galerkin methods with the ultra-weak formulation}
\seclab{abstractUweak}

The name of the discontinuous Petrov-Galerkin method refers to the fact that the method is a Petrov-Galerkin method, and that the test functions are specified to be discontinuous across element boundaries. There is no specification of the regularity of the trial space, and we stress that the idea of DPG is not inherently tied to a single variational formulation \cite{Bui-ThanhDemkowiczGhattas11a}. Additionally, Cohen, Dahmen and Welper simultaneously extended the minimum residual concept behind DPG to general variational settings and avoid the use of discontinuous test functions by formulating the minimum residual method as a saddle-point problem \cite{DahmenVariationalStabilization}.

In most of the DPG literature, however, the discontinuous Petrov-Galerkin method refers to the combination of the concept of locally computable optimal test functions introduced in Section \secref{optimalTest} with the so-called ``ultra-weak formulation" \cite{DPG1,DPG2,DPG3,DPG4,DPGElas,DBLP:journals/procedia/NiemiCC11}. Unlike the previous two sections in which we studied the general equation \eqnref{variationEq} given by abstract bilinear and linear forms, we now consider a concrete instance of \eqnref{variationEq} resulting from an ultra-weak formulation for an abstract first-order system of PDEs $Au = f$. Additionally, from this section onwards, we will refer to DPG as the pairing of the ultra-weak variational formulation with the concept of locally computable optimal test functions.

We begin by partitioning the domain of interest $\Omega$ into $\Nel$ non-overlapping elements $K_j, j = 1,\hdots,\Nel$ such that $\Oh = \cup_{j=1}^\Nel K_j$ and $\overline{\Omega} = \overline{\Omega}_h$. Here, $h$ is defined as $h= \max_{j\in \LRc{1,\hdots,\Nel}}\text{diam}\LRp{K_j}$.  We denote the mesh ``skeleton" by $\Gh = \cup_{j=1}^\Nel \partial K_j$; the set of all faces/edges $e$, each of which come with a normal vector ${n}_e$. The internal skeleton is then defined as $\Gamma^0_h = \Gh \setminus \partial \Omega$. If a face/edge $e \in \Gh$ is the intersection of $\partial K_i$ and $\partial K_j$, $i \ne j$, we define the following jumps:
\[
\jump{v} = \text{sgn} \LRp{{n}^-}v^- + \text{sgn} \LRp{{n}^+}v^+, \quad
\jump{\tau \cdot n} = {n}^-\cdot \tau^- + {n}^+\cdot\tau^+,
\]
where
\[
\text{sgn}\LRp{{n}^{\pm}} =
\left\{
\begin{array}{ll}
1 & \text{if } {n}^{\pm} = {n}_e \\
-1 & \text{if } {n}^\pm = -{n}_e
\end{array}
\right..
\]
For $e$ belonging to the domain boundary $\partial \Omega$, we define
\[
\jump{v} = v, \quad
\jump{\tau \cdot n} = {n}_e\cdot \tau.
\]
Note that we allow arbitrariness in assigning ``$-$'' and ``$+$'' quantities to the adjacent elements $K_i$ and $K_j$.
%For the rest of the paper, we will use the same notation for both a function and its trace (if it is well-defined) when there is no ambiguity.

The ultra-weak formulation for $Au = f$ on $\Oh$, ignoring boundary
conditions for now, reads
\begin{equation}
\eqnlab{uweak}
b\left(\left(u, \widehat{u}\right),v\right) \coloneqq \langle \widehat{u}, \jump{v}
\rangle_{\Gh} - (u,A_h^*v)_{\Oh}= \LRp{f,v}_{\Oh},
\end{equation}
where we have denoted $\LRa{\cdot,\cdot}_{\Gh}$ as the duality
pairing on $\Gh$, $\LRp{\cdot,\cdot}_{\Oh}$ the $L^2$-inner
product over $\Oh$, and $A_h^*$ the formal adjoint resulting from
element-wise integration by parts.\footnote{We note that $\jump{v}$ refers to the jump of the trace of the test function across an edge; for example, for vector valued test functions $\tau$, $\jump{\tau}$ corresponds to the jump of the normal component. We will continue to use this notation in following chapters.}
Occasionally, for simplicity in
writing, we will ignore the subscripts in the duality pairing and
$L^2$-inner product if they are $\Gh$ and $\Oh$. Both the
inner product and formal adjoint are understood to be taken
element-wise. Using the ultra-weak formulation, the regularity
requirement on solution variable $u$ is relaxed, that is, $u$ is now
square integrable for the ultra-weak formulation \eqnref{uweak} to be
meaningful, instead of being (weakly) differentiable.  The trade-off
is that $u$ does not admit a trace on $\Gh$ even though it did
originally. Consequently, we need to introduce an additional new
``trace'' variable $\widehat{u}$ in \eqnref{uweak}, that is defined only on
$\Gh$.

@@

%\subsection{Optimal and quasi-optimal test norms}
\section{A canonical energy norm pairing for ultra-weak formulation}

@@

Using variants of the graph test norm, numerical results show that the DPG method appears to provide a ``pollution-free" method without phase error for the Helmholtz equation \cite{DPG4}, and analysis of the pollution-free nature of DPG is currently under investigation. Similar results have also been obtained in the context of elasticity \cite{DPGElas} and the linear Stokes equations \cite{Camellia}. On the theoretical side, the graph test norm has been shown to yield a well-posed DPG methodology for the Poisson and convection-diffusion equations \cite{analysisDPG}. More recently, this theory has been generalized to show the well-posedness of DPG for the large class of PDEs of Friedrichs' type \cite{Bui-ThanhDemkowiczGhattas11b}.
